\Large Clustering Vehicles based on Trips Identified from Automatic \\ Number Plate Recognition Camera Scans

### 2018-02-26 classification.tex

\begin{table*}[t]
\centering
\begin{tabular}{c c c c c c c c c c c}
  \hline
 \thead{Total\\Trips} & \thead{Average\\Trips} & \thead{Average\\Degree} & \thead{Average\\Sightings} & \thead{Average\\Distinct\\Origins} & \thead{Average\\Distinct\\Destinations} & \thead{Average\\Distinct\\Routes} & \thead{Average\\First\\Hour} & \thead{Average\\Last\\Hour} & \thead{Average\\Hour\\Difference} & \thead{Average\\Rest\\Time} \\
  \hline
41 & 3.42 & 1.25 & 5.75 & 2.75 & 1.00 & 3.00 & 15.33 & 19.25 & 3.80 & 3.70 \\
3 & 1.50 & 1.00 & 1.50 & 1.50 & 0.00 & 1.50 & 14.00 & 14.50 & 0.73 & 0.73 \\
7 & 2.33 & 1.33 & 3.33 & 2.33 & 0.67 & 2.33 & 11.00 & 13.00 & 2.44 & 2.41 \\
12 & 2.40 & 1.10 & 3.60 & 2.40 & 0.60 & 2.40 & 14.40 & 16.40 & 2.00 & 1.95 \\
   \hline
\end{tabular}
\caption{Sample of extracted features from trips taken from 15 weekdays of number plate data.}
\vspace{-0.2cm}
\label{t:features}
\end{table*}

@@

Due to distinctly different traffic behaviour during weekends, we only consider trips occurring during weekdays. Therefore, we used all number plate data collected between the 6th and 24th of February and excluded data from the 2 weekends in-between. Furthermore, as mentioned in section~\ref{s.trips}, we used fixed empirical values for $\Tau$. Table~\ref{t:trips-tau} displays the number of trips, average trip degree and the proportion of trips of degree one, for varying values of $\Tau$. The effect of incrementing $\Tau$ on resulting trips is clear: increased trip degrees and decreased trips containing just a single sighting. On the other hand, we did not set a value for $\tau$, but instead handled implausible trips by filtering all sightings with confidence below 85\%. Duplicates were filtered after identifying consecutive sightings by the same camera occurring within the same trip. Clock synchronisation errors were provided in milliseconds with none exceeding 5 seconds. These were hence ignored.

@@

Table~\ref{t:features} depicts a sample of the resulting features vector. A total of 1,034,107 distinct vehicles were detected. However, because there is a high percentage of trips containing a single sighting, some of these features were highly correlated. We therefore, chose to remove three of the features represented in table~\ref{t:features}: \emph{Average Sightings}, \emph{Average Distinct Routes} and \emph{Average Hour Difference}, to avoid the obfuscation of the natural clustering ~\cite{Kmeans}. Furthermore we considered that a trip of degree one has no destination (which explains values of average distinct destination below one) and we filtered all instances of vehicles where the total number of trips is lower than 3, resulting in 642,006 unique vehicles.

Clustering of vehicles was performed using the Hartigan and Wong \emph{k-means} algorithm, for each value of $\Tau$. The number of clusters \emph{k} was varied between 2 and 8 and executed with a maximum of 200 iterations and 100 different starting states of the algorithm. The Calinski-Harabasz criterion is used to determine the best value of $k$, the one that minimises the within-cluster and between-cluster errors and provides the more natural clusters ~\cite{Kmeans}. The results of trip clustering are presented and discussed in the next section.

@@

Table~\ref{t:tau_comparison} provides a summary of multiple runs of \emph{k-means}. For each value of $\Tau$, the optimal number of clusters is selected by picking the value of $k$ that maximises the Calinski-Harabasz criterion. Furthermore, the measures of inter-cluster (betweeness) and intra-cluster (withiness) are depicted relative to the corresponding total sum of squares (total betweeness and total withiness). It is noteworthy that the best value of $k$ increases inversely to $\Tau$. Although a higher value of $k$ seems to suggest that trips with smaller values of $\Tau$ are better able to capture the variance of trip data, we have to consider that varying $\Tau$ affects the average trip degree in the same direction whilst affecting the total number of trips in the opposite direction (table~\ref{t:trips-tau}).

@@

On the other hand, we expected to observe a group representing home to work commuters with the first trip of the day starting approximately at eight in the morning and a second trip terminating between five and six in the evening. Although we observe one or two groups with those characteristics, these contain a higher average of trips per day than expected, which may represent for example work to school trips. However, we observe at least one big group of trips occurring mostly during lunch hour. A big contributing factor however is the fact that a high proportion of trips contains only a single sighting (table~\ref{t:trips-tau}). It may be the case that many of these drivers choose routes other than those going through ANPR cameras.

To improve the interpretability of these results, we've gain access to a database that outputs the vehicle's weight, wheelplane, make and category from the corresponding plate number. Broadly speaking, vehicle categories are defined relative to the transport of persons or goods, and sub-categories are given according to maximum allowed mass. With this information we sorted vehicles in the ANPR database into one of the following classes: \textit{Bus, Car, Motorcycle, Truck, Van}. Unfortunately, we were unable to identify taxis and therefore were unable to separate private cars from taxis. Table~\ref{t:} shows the mean relative frequency of these classes, and the corresponding deviation for the clusters shown in table~\ref{t:kmeans_centers_300}. Only the clusters showing more notable deviations from the mean are shown. Cluster 5, in particular is mainly composed by trucks. Cluster 6, sees an increase of frequency of cars which may represent taxis. Vans are seen in larger proportions in Cluster 7 and 3. These results provide some evidence in favor of some of our assumptions above and demonstrate that whilst ANPR data is able to capture some of the travel patterns, a clear and robust assessment of travel patterns is still not possible. Furthermore, it is not clear whether this is due to limited coverage of the road network or to a lack of methods that model, quantify and address the uncertainty of trips identified from ANPR data.

@@

\begin{table}[t]
\centering
\tabcolsep=0.17cm
\begin{tabular}{c c c c c}
  \hline
Tau & Best $k$ & Betweenss & \thead{Average\\Withinss} & \thead{Calinski-\\Harabasz} \\
  \hline
5 min &   8 & 0.923 & 0.125 & 1,116,962 \\
  7.5 min &   7 & 0.904 & 0.143 & 1,003,744 \\
  10 min&   7 & 0.896 & 0.143 &   911,044 \\
  15 min &   6 & 0.865 & 0.167 &   794,557 \\
  20 min &   4 & 0.786 & 0.250 &   754,241 \\
  30 min &   3 & 0.710 & 0.333 &   743,001 \\
   \hline
\end{tabular}
\caption{\emph{k-means} performance for several values of $\Tau$.}
\label{t:tau_comparison}
\end{table}

>>>>

@@

Due to distinctly different traffic behaviour during weekends, we only consider trips occurring during weekdays. Therefore, we used all number plate data collected between the 6th and 24th of February and excluded data from the 2 weekends in-between. Furthermore, as mentioned in section~\ref{s.trips}, we used fixed empirical values for $\Tau$. Table~\ref{t:trips-tau} displays the number of trips, average trip degree and the proportion of trips of degree one, for varying values of $\Tau$. The effect of incrementing $\Tau$ on resulting trips is clear: increased trip degrees and decreased trips containing just a single sighting. On the other hand, we did not set a value for $\tau$, but instead handled implausible trips by filtering out all sightings with confidence below 85\%. Duplicates were filtered after identifying consecutive sightings by the same camera occurring within the same trip. Clock synchronisation errors were provided in milliseconds with none exceeding 5 seconds. These were hence ignored.

\begin{table}[t]
\centering
\tabcolsep=0.25cm
\begin{tabular}{c c c c}
  \hline
$\Tau$ & Trips & \thead{Average\\Degree} & \thead{Proportion of Trips \\ Degree 1} \\
  \hline
5 min & 13,603,759 & 1.46 & 0.70 \\
7.5 min & 12,394,709 & 1.60 & 0.64 \\
10 min & 11,690,791 & 1.69 & 0.61 \\
15 min & 10,823,333 & 1.83 & 0.57 \\
20 min & 10,305,860 & 1.92 & 0.54 \\
30 min &  9,653,499 & 2.04 & 0.52 \\
   \hline
\end{tabular}
\caption{Overview of trip data for varying values of $\Tau$.}
\label{t:trips-tau}
\vspace{-0.5cm}
\end{table}

@@

Table~\ref{t:features} depicts a sample of the resulting features vector. A total of 1,034,107 distinct vehicles were detected. However, because there is a high percentage of trips containing a single sighting, some of these features were highly correlated. We therefore, chose to remove three of the features represented in Table~\ref{t:features}: \emph{Average Sightings}, \emph{Average Distinct Routes} and \emph{Average Hour Difference}, to avoid the obfuscation of the natural clustering ~\cite{Kmeans}. Furthermore we considered that a trip of degree one has no destination (which explains values of average distinct destination below one) and we filtered all instances of vehicles where the total number of trips is lower than 3, resulting in 642,006 unique vehicles.

Clustering of vehicles was performed using the Hartigan and Wong \emph{k-means} algorithm, for each value of $\Tau$. The number of clusters \emph{k} was varied between 2 and 8 and executed with a maximum of 200 iterations and 100 different starting states of the algorithm. The Calinski-Harabasz criterion is used to determine the best value of $k$. It minimises and maximises the within-cluster and between-cluster sum of squares respectively, resulting in more natural clusters ~\cite{Kmeans}. The results of trip clustering are presented and discussed in the next section.

@@

Table~\ref{t:tau_comparison} provides a summary of multiple runs of \emph{k-means}. For each value of $\Tau$, the optimal number of clusters is selected by picking the value of $k$ that maximises the Calinski-Harabasz criterion. Furthermore, the measures of inter-cluster (betweeness) and intra-cluster (withiness) are depicted relative to the corresponding total sum of squares (total betweeness and total withiness). It is noteworthy that the best value of $k$ increases inversely to $\Tau$. Although a higher value of $k$ seems to suggest that trips with smaller values of $\Tau$ are better able to capture the variance of trip data, we have to consider that varying $\Tau$ affects the average trip degree in the same direction whilst affecting the total number of trips in the opposite direction (Table~\ref{t:trips-tau}).

@@

On the other hand, we expected to observe a group representing home to work commuters with the first trip of the day starting approximately at eight in the morning and a second trip terminating between five and six in the evening. Although we observe one or two groups with those characteristics, these contain a higher average of trips per day than expected, which may represent for example work to school trips. However, we observe at least one big group of trips occurring mostly during lunch hour. A big contributing factor however is the fact that a high proportion of trips contains only a single sighting (Table~\ref{t:trips-tau}). It may be the case that many of these drivers choose routes other than those going through ANPR cameras.

To improve the interpretability of these results, we would like to gain access to a governmental database that provides vehicles' weight, wheelplane, make and category from the corresponding plate number. Broadly speaking, vehicle categories are defined relative to the transport of persons or goods, and sub-categories are given according to maximum allowed mass. With this information it would be possible to sort vehicles in the ANPR database into one of the following classes: \textit{Bus, Car, Motorcycle, Truck, Van}. Even though taxis cannot be identified in this way, taxi companies could be contacted directly to obtain this data. However, such databases are available for law enforcement, but not to the public for ethical and privacy reasons. Additionally, this would require us to de-anonymize the hashed number plates. Nevertheless, these results demonstrate that whilst ANPR data is able to capture some of the travel patterns, a clearer and more robust assessment of travel patterns is needed. Furthermore, it is not clear whether this is due to limited coverage of the ANPR cameras in the road network or of the methodology used.

@@

\begin{table}[t]
\centering
\tabcolsep=0.17cm
\begin{tabular}{c c c c c}
  \hline
$\Tau$ & Best $k$ & Betweenss & \thead{Average\\Withinss} & \thead{Calinski-\\Harabasz} \\
  \hline
5 min &   8 & 0.923 & 0.125 & \textbf{1,116,962} \\
  7.5 min &   7 & 0.904 & 0.143 & 1,003,744 \\
  10 min&   7 & 0.896 & 0.143 &   911,044 \\
  15 min &   6 & 0.865 & 0.167 &   794,557 \\
  20 min &   4 & 0.786 & 0.250 &   754,241 \\
  30 min &   3 & 0.710 & 0.333 &   743,001 \\
   \hline
\end{tabular}
\caption{\emph{k-means} performance for several values of $\Tau$.}
\label{t:tau_comparison}
\end{table}

### 2018-02-26 intro.tex
Made some text changes and some re-placement of tables/figures: discussion/errors in plate scanning

In a utopian model we would have full disclosure of all journeys made by all vehicles on the road infrastructure. However, this has numerous ethical and technical issues. From an ethical standpoint should we be allowed to know where all vehicles are at any given point in time. From a technical point of view, although every vehicle could be fitted with a GPS tracker -- costly in its own right -- there would still exist the issue of how we would collect and stream all of this data for future processing. Alternatively one can view the problem the other way around and rather than tracking individual vehicles look at collecting information by observing vehicles passing points within the road networks. A prime example of this approach are Automatic Number Plate Recognition (ANPR) cameras. These cameras are a combination of digital camera coupled with Artificial Intelligence to identify number plates within the image and convert these into strings of characters. ANPR cameras are normally fixed in location\footnote{Although cameras can be in a vehicle and moved from location to location.} able to view all vehicles passing that location.

For ANPR the problem now becomes that of recovering as much information about vehicle's journey as possible from the limited number of observations. ANPR cameras are normally located on major roads and interchanges, however, this only covers a tiny fraction of the road network. We can, though, estimate routes between cameras by understanding the distances between cameras and the most ``sensible'' routes between them. This allows us, given a set of ANPR sightings of the same vehicle, to produce a ``most likely'' route for that journey. It should be noted that we cannot determine the actual start and end of the journey as these will happen in areas not covered by ANPR. It should also be noted that for ethical reasons it is not normal to obtain actual number plates, but rather the hash of these. Though, for most situations this will suffice.

Once we have a set of sightings of a vehicle using ANPR, we now need to convert these into actual journeys. The first requirement is to identify individual journeys. Although this can't be done with certainty we can apply general rules to distinguish one journey from the next. For example if two sightings are made from ANPR cameras which are connectable by a ``sensible'' route\footnote{Here ``sensible'' implies that a route between cameras A and B would not need to go through a third camera C.} in a time interval which is ``sensible'' then these can be determined to be part of the same journey. However, if the timings between two sightings is significantly longer than what would be expected then this would imply that the vehicle stopped between these two cameras and that the later sighting is part of a new journey.

The process of journey identification needs to be performed on dirty data which contains numerous impurities which need to be handled. These include:

\begin{itemize}
	\item {\bf Number plate miss-reads:} Although ANPR cameras have accuracies of around 99, miss-reads are possible. This can lead to sightings being missed or vehicles being wrongly sighted in locations.
	\item {\bf Timing errors:} The time-stamps of sightings could be erroneous. The minor side of this is implausible journey times, though, more seriously, this can lead to reordering the set of cameras on a particular journey.
	\item {\bf Clones number plates:} For various reasons a number plate may be cloned and used on a different vehicle. This can lead to impossible journeys and journeys that the real vehicle did not make.
\end{itemize}

@@

The rest of this paper is presented as follows. In Section \ref{s.related} we discuss related work. Section \ref{s.ncl} we presents the ANPR data for the Newcastle area. Our process for identifying individual journeys is presented in Section \ref{s.trips} while Section \ref{s.classification} presents our classification approach. We present results in Section \ref{s.results} before offering conclusions and future directions in Section \ref{s.conclusions}.

@@

\begin{figure}[t]
\centering
\includegraphics[width=.90\linewidth]{ANPR-overview.pdf}
\caption{Overview of an ANPR system.}
\label{fig:anpr-overview}
\vspace{-0.6cm}
\end{figure}

The main use of ANPR data for the UTMC is estimating average journey times for selected or sensitive links in the road network. Furthermore, several authors have extensively researched how to use number plate data as an extension to link counts for estimating origin-destination matrices and link flows~\cite{Castillo2010, Castillo2008, Hazelton2012}. However, very few works have focused on analysing individual or collective travel patterns from number plate data, particularly across extended periods of time. Moreover, there is no consistent conceptual and analytical framework for transforming number plate data into a historical sequence of trips for each vehicle. Finally, we believe that trip data, properly identified from number plate data, has the potential to unlock a number of new applications for urban traffic control and law enforcement. Thus, in section~\ref{s.trips} we present a conceptual methodology for grouping multiple camera observations of the same vehicle into one or several trips of that vehicle.

Determining the distribution of travel modes is one of the fundamental steps in the four-stage model: an essential traffic modelling methodology for transportation planning~\cite{FourStepModel}. Previous works have used trip information derived from different sources of data to identify travel mode or purpose of trip. More notably, survey data, floating car data and mobile phone data have been used~\cite{ODMobileData, ClusteringGPS}.  Although number plate data has been used in~\cite{Clustering} to identify different categories of trips, the authors do not differentiate between private or public travel modes and focus instead on categorising trips by time of occurrence. Hence, in section~\ref{s.classification} we apply the \emph{k-means} clustering algorithm to derived trip data and based on the results, we discuss the limitations of proposed methods and that of ANPR data.

>>>>

In a utopian model we would have full disclosure of all journeys made by all vehicles on the road infrastructure. However, this has numerous ethical and technical issues. From an ethical standpoint should we be allowed to know where all vehicles are at any given point in time? From a technical point of view, although every vehicle could be fitted with a GPS tracker -- costly in its own right -- there would still exist the issue of how we would collect and stream all of this data for future processing. Alternatively one can view the problem the other way around and rather than tracking individual vehicles look at collecting information by observing vehicles passing points within the road network. A prime example of this approach are Automatic Number Plate Recognition (ANPR) cameras. These cameras are a combination of digital camera coupled with Artificial Intelligence to identify number plates within the image and convert these into strings of characters. ANPR cameras are normally fixed in location\footnote{Although cameras can be in a vehicle and moved from location to location.} able to view all vehicles passing that location.

For ANPR the problem now becomes that of recovering as much information about a vehicle's journey as possible from the limited number of observations. ANPR cameras are normally located on major roads and interchanges, however, this only covers a tiny fraction of the road network. We can, however, estimate routes between cameras by understanding the distances between cameras and the most ``sensible'' routes between them. This allows us, given a set of ANPR sightings of the same vehicle, to produce a ``most likely'' route for that journey. It should be noted that we cannot determine the actual start and end of the journey as these will happen in areas not covered by ANPR. It should also be noted that for ethical reasons it is not normal to obtain actual number plates, but rather the hash of these. Though, for most situations this will suffice.

Once we have a set of sightings of a vehicle using ANPR, we now need to convert these into actual journeys. The first requirement is to separate the stream of sightings of a vehicle into individual journeys. Although this can't be done with certainty we can apply general rules to distinguish one journey from the next. For example if two sightings are made from ANPR cameras which are connectable by a ``sensible'' route\footnote{Here ``sensible'' implies that a route between cameras A and B would not need to go through a third camera C.} in a time interval which is ``sensible'' then these can be determined to be part of the same journey. However, if the timings between two sightings is significantly longer than what would be expected then this would imply that the vehicle stopped between these two cameras and that the later sighting is part of a new journey. The process of journey identification needs to be performed on dirty data which contains numerous impurities which need to be handled. These include:

\setlist{nosep}
\begin{itemize}
	\item {\bf Number plate miss-reads:} Although ANPR cameras have accuracies of around 99\%, miss-reads are possible. This can lead to sightings being missed or vehicles being wrongly sighted in locations.
	\vspace{0.1cm}
	\item {\bf Timing errors:} The time-stamps of sightings could be erroneous. The minor side of this is implausible journey times, though, more seriously, this can lead to reordering the set of cameras on a particular journey.
	\vspace{0.1cm}
	\item {\bf Cloned number plates:} For various reasons a number plate may be cloned and used on a different vehicle. This can lead to impossible journeys and journeys that the real vehicle did not make.
\end{itemize}

@@

The rest of this paper is presented as follows. In Section~\ref{s.related} we discuss related work. Section~\ref{s.ncl} presents the ANPR data for the Newcastle area. Our process for identifying individual journeys is defined in Section~\ref{s.trips}, while methods for addressing issues resulting from poor camera performance are described in Sections~\ref{s.duplicates} and~\ref{s.errors}. Section~\ref{s.classification} outlines our clustering approach. We report results in Section~\ref{s.results} before offering conclusions and future directions in Section~\ref{s.conclusions}.

@@

The main use of ANPR data for an Urban Traffic Management and Control Centre (UTMC) is estimating average journey times for selected or sensitive links in the road network. Furthermore, several authors, notably Enrique Castillo et al.\ and Andrew Hazelton et al.\, have extensively researched how to use number plate data as an extension to link counts for estimating origin-destination matrices and link flows~\cite{Castillo2010, Castillo2008, Hazelton2012}. However, very few works have focused on analysing individual or collective travel patterns from number plate data, particularly across extended periods of time. Moreover, there is no consistent conceptual and analytical framework for transforming number plate data into a historical sequence of trips for each vehicle. Finally, we believe that trip data, properly identified from number plate data, has the potential to unlock a number of new applications for urban traffic control and law enforcement. Thus, in section~\ref{s.trips} we present a conceptual methodology for grouping multiple camera observations of the same vehicle into one or several trips of that vehicle.

Determining the distribution of travel modes is third of the four fundamental steps in the four-stage model: trip generation, trip distribution, modal split and traffic assignment. The four-stage model is the most widely used traffic modelling methodology for transportation planning~\cite{FourStepModel}. Previous works have used trip information derived from different sources of data to identify travel mode or purpose of trip. More notably, survey data, floating car data and mobile phone data have been used~\cite{ODMobileData, ClusteringGPS}.  Although number plate data has been used by Chen et al.\ (2017)~\cite{Clustering} to identify different categories of trips, the authors do not differentiate between private or public travel modes and focus instead on categorising trips by time of occurrence. Hence, in section~\ref{s.classification} we apply the \emph{k-means} clustering algorithm to derived trip data and based on the results, we discuss the limitations of proposed methods and that of ANPR data.
